---
title: "BCB420 Assignment 2 - Differential Gene expression and Preliminary ORA"
author: Arianne Beauregard
output: 
  html_notebook:
    toc: true
    depth: 2
bibliography: references.bib
---

# Install and load required packages
```{r, echo = T, results='hide'}
# all required packages
bioc <- c("ComplexHeatmap", "edgeR", "limma")
cran <- c("knitr", "kableExtra", "circlize", "ggplot2", "dplyr", "gprofiler2")

# install BiocManager if needed
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# install bioconductor packages if necessary
for (package in bioc) {
  if (!require(package, character.only = TRUE)) {
    BiocManager::install(package)
    require(package)
  }
}

# install cran packages if necessary
for(package in cran) {
  if(!require(package, character.only = TRUE)){
    install.packages(package)
    require(package)
  }
}
```

---

# Introduction on Data
The dataset is from a study of lipid remodeling in response to methionine stress in triple-negative breast cancer (TNBC) cells (Borrego et al., 2021). The dataset contains RNA profiles of two TNBC cell lines, one methionine-sensitive cell line (MDA-MBA-468) its methionine-insensitive derivative (MDA-MB-468res-R8), responding to methionine stress. The data was downloaded from GEO Accession GSE155955.

In Assignment 1, the data was preprocessed and normalized. Gene identifiers were mapped to their HUGO Symbols. The data was normalized using Trimmed Mean of M-values (TMM). 

We will download the normalized data from Assignment 1 and take a look at the data.
```{r}
normalized_count_data <- readRDS("../data/normalized_count_data.rds")

kable(normalized_count_data[1:5,1:5], type="html", caption = "Table 1. A brief look at the data") %>%
  kable_styling()
```

---

# Differential Gene Expression
## Determining factors
We first need to determine which factors to account for in our analysis. Let's revisit the sample information. 

There are two different cell lines (MDA-MB-468 and MDA-MB-468res-R8), measured at 4 different timepoints in hours (h) (0h, 0.5h, 2h, 12h) during methionine stress. Methionine stress was added by providing homocysteine instead of methionine in the cell lines' growth media. There are 3 replicates for each cell line. 
```{r}
# define groups
samples <- data.frame(lapply(colnames(normalized_count_data), 
        FUN=function(x){unlist(strsplit(x, 
                        split = "_"))}))
colnames(samples) <- colnames(normalized_count_data)
rownames(samples) <- c("cell.line","timepoint", "replicate")
samples <- data.frame(t(samples))
kable(samples, caption = "Table 2. Sample Groups", col.names = c("Cell Line", "Timepoint (min)", "Replicate")) %>%
  kable_styling()
```

We will visualize the factors of the samples in an MDS plot (Figure 1).
```{r, fig.cap = "Multidimensional scaling (MDS) plot of samples"}
colors <- c("blue", "red", "green4", "purple")
colors <- colors[factor(samples$timepoint)]
symbols <- c(1, 2)
symbols <- symbols[factor(samples$cell.line)]

limma::plotMDS(normalized_count_data,
               col = colors, pch = symbols, ylim = c(-1, 2.5), cex = 1.5)
title("MDS Plot of Samples")
legend("topright", title = "Timepoint (min)", fill = unique(colors), legend = unique(factor(samples$timepoint)), cex = 0.6, ncol = 4, bty = 'n', border = F)
legend(x = 1.25, y = 2.2, title = "Cell Line", pch=unique(symbols),legend=factor(unique(factor(samples$cell.line))), cex = 0.6, ncol = 2, bty = 'n', border = F)
```
__Figure 1. Multidimensional scaling (MDS) plot of samples.__ The cluster at the bottom left is are samples of MDA-MB-468 cell line at times 0 min, 30 min and 120 min. The cluster at the bottom right is samples of  MDA-MB-468res-R8 cell line at times 0 min, 30 min and 120 min. The cluster at the top left is samples of MDA-MB-468 cell line at time 720 min. The cluster at the top right is samples MDA-MB-468res-R8 cell line at time 720 min.


<br>

From the MDS plot, we can see 4 clear clusters. The clustering is based on both the cell type and timepoint, therefore our model design should account for both these factors. 

---

## Differential expression analysis using edgeR
We will do two separate differential expression analyses; one for cell line MDA-MB-468, and the other for cell line MDA-MB-468res-R8. In each analysis, the cell line 12 hours after of methionine stress (with timepoint == 720) will be compared with the others. The edgeR package will be used as it is designed for differential expression analysis of RNA-seq expression profiles with biological replicates [@edgeR].


### Analysis for cell line MDA-MB-468
First, we prepare the count data and sample information for samples of cell line MDA-MB-468, The model design is based on timepoint. We prepare a DGEList of the normalized count data and the groups, and then fit a quasi-likelihood negative binomial generalized log-linear model to our data and calculate the differential expression. We compare samples at timepoint 720 min vs the other samples.

```{r}
counts_MB468 <- normalized_count_data[, grepl("MB468", colnames(normalized_count_data))]
samples_MB468 <- samples[colnames(counts_MB468),]


model_design <- model.matrix(~ samples_MB468$timepoint)
# DGEList 
d <- DGEList(counts = counts_MB468, group = samples_MB468$timepoint)
# Estimate Dispersion with model design.
d <- estimateDisp(d, model_design)
# Fit the model
fit <- glmQLFit(d, model_design)
qlf.results <- glmQLFTest(fit, coef = "samples_MB468$timepoint720")
# take a look at results
kable(topTags(qlf.results)$table, type="html", caption = "Table 3. Results from fitting negative binomial generalized log-linear model to normalized count data of MDA-MB-468 samples") %>%
  kable_styling()
```

We will look at the results to see how many genes pass the threshold p-value < 0.05; if the p-value is lower than 0.05, we consider the gene to be significantly differentially expressed. We also look at how many genes pass the correction. The correction method used is Benjamini-Hochberg (BH) method, which controls the false discovery rate (FDR).
```{r}
# all results
qlf_output_hits_MB468 <- topTags(qlf.results,sort.by = "PValue",
                           n = nrow(counts_MB468))$table

# pass the threshold p-value < 0.05
threshold_pval <- length(which(qlf_output_hits_MB468$PValue < 0.05))
print(paste("Number of genes that pass the threshold p-value < 0.05:", threshold_pval))

# pass correction
correction <- length(which(qlf_output_hits_MB468$FDR < 0.05))
print(paste("Number of genes that pass correction:", correction))
```

Next, we will see how many genes were significantly differentially expressed. Of the genes that pass the p-value threshold of 0.05, if the log fold change is greater than 0, the gene is upregulated, and if it is less than 0, the gene is downregulated.
```{r}
# adding expression column to record if gene is upregulated or downregulated
qlf_output_hits_MB468 <- qlf_output_hits_MB468 %>% 
  mutate(
    expression = case_when(PValue < 0.05  & logFC > 0 ~ "upregulated",
                           PValue < 0.05  & logFC < 0 ~ "downregulated",
                           TRUE ~ "unchanged")
    )

# upregulated
print(paste("Number of upregulated genes (MDA-MB-468):",
            length(which(qlf_output_hits_MB468$expression == "upregulated"))))

# downregulated
print(paste("Number of upregulated genes (MDA-MB-468):",
            length(which(qlf_output_hits_MB468$expression == "downregulated"))))
```


We will visualize this in a volcano plot (Figure 2).
```{r, fig.cap = "Volcano Plot of differential expression in MDA-MB-468 samples"}
# volcano plot
volcano_plot <- ggplot(qlf_output_hits_MB468, aes(logFC, -log(PValue,10))) +
  geom_point(aes(color = expression), size = 2/5) +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"PValue")) +
  ggtitle("Plot of differential expression in MDA-MB-468 samples") + 
  scale_color_manual(values = c("blue", "dimgrey", "red")) +
  guides(colour = guide_legend(override.aes = list(size=1.5))) 
volcano_plot
```
__Figure 2. Volcano Plot of differential expression in MDA-MB-468 samples.__ The points on the plot represent genes. Red points are upregulated, blue points are downregulated, and grey points are unchanged. 


### Analysis for cell line MDA-MB-468res-R8
The same process is carried out for samples of cell line MDA-MB-468res-R8.
```{r}
counts_R8 <- normalized_count_data[, grepl("R8", colnames(normalized_count_data))]
samples_R8 <- samples[colnames(counts_R8),]


model_design <- model.matrix(~ samples_R8$timepoint)
# DGEList 
d <- DGEList(counts = counts_R8, group = samples_R8$timepoint)
# Estimate Dispersion with model design.
d <- estimateDisp(d, model_design)
# Fit the model
fit <- glmQLFit(d, model_design)
qlf.results <- glmQLFTest(fit, coef = "samples_R8$timepoint720")
# take a look at results
kable(topTags(qlf.results)$table, type="html", caption = "Table 4. Results from fitting negative binomial generalized log-linear model to normalized count data of MDA-MB-468res-R8 samples") %>%
  kable_styling()
```

We will look at the results to see how many genes pass the threshold p-value < 0.05, and how many genes pass the correction. The same methods for the previous analysis are applied here.
```{r}
# all results
qlf_output_hits_R8 <- topTags(qlf.results,sort.by = "PValue",
                           n = nrow(counts_R8))$table
# pass the threshold p-value < 0.05
threshold_pval <- length(which(qlf_output_hits_R8$PValue < 0.05))
print(paste("Number of genes that pass the threshold p-value < 0.05:", threshold_pval))

# pass correction
correction <- length(which(qlf_output_hits_R8$FDR < 0.05))
print(paste("Number of genes that pass correction:", correction))
```

Next, we will see how many genes were significantly differentially expressed. 
```{r}
# adding expression column to record if gene is upregulated, downregulated or unchanged
qlf_output_hits_R8 <- qlf_output_hits_R8 %>% 
  mutate(
    expression = case_when(PValue < 0.05  & logFC > 0 ~ "upregulated",
                           PValue < 0.05  & logFC < 0 ~ "downregulated",
                           TRUE ~ "unchanged")
    )

# upregulated
print(paste("Number of upregulated genes (MDA-MB-468res-R8):",
            length(which(qlf_output_hits_R8$expression == "upregulated"))))

# downregulated
print(paste("Number of upregulated genes (MDA-MB-468res-R8):",
            length(which(qlf_output_hits_R8$expression == "downregulated"))))
```


We will visualize this in a volcano plot (Figure 3).
```{r, fig.cap = "Volcano Plot of differential expression in MDA-MB-468res-R8 samples"}
# volcano plot
volcano_plot <- ggplot(qlf_output_hits_MB468, aes(logFC, -log(PValue,10))) +
  geom_point(aes(color = expression), size = 2/5) +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"PValue")) +
  ggtitle("Plot of differential expression in MDA-MB-468res-R8 samples") + 
  scale_color_manual(values = c("blue", "dimgrey", "red")) +
  guides(colour = guide_legend(override.aes = list(size=1.5))) 
volcano_plot
```
__Figure 3. Volcano Plot of differential expression in MDA-MB-468res-R8 samples.__ The points on the plot represent genes. Red points are upregulated, blue points are downregulated, and grey points are unchanged. 

## Heatmap of top hits
We will visualize all our top hits with a heatmap (Figure 4). We will create the heatmap using the data from the differentially expressed genes.
```{r, fig.cap = "Heatmap of differentially expressed genes"}
# getting all differentially expressed genes (not unchaged)
mb468_deg <- row.names(qlf_output_hits_MB468)[qlf_output_hits_MB468$expression != "unchanged"]
r8_deg <- row.names(qlf_output_hits_R8)[qlf_output_hits_R8$expression != "unchanged"]

rows <- union(mb468_deg, r8_deg)

top_hits <- normalized_count_data[rows, ]

heatmap_matrix <- t(scale(t(top_hits)))
if(min(heatmap_matrix) == 0){
    heatmap_col = colorRamp2(c( 0, max(heatmap_matrix)), 
                      c( "white", "red"))
  } else {
    heatmap_col = colorRamp2(c(min(heatmap_matrix), 0,
        max(heatmap_matrix)), c("blue", "white", "red"))
  }
tophits_heatmap <- Heatmap(as.matrix(heatmap_matrix),
      show_row_dend = TRUE,show_column_dend = TRUE, 
      col=heatmap_col,show_column_names = TRUE, 
      show_row_names = FALSE, show_heatmap_legend = TRUE)

tophits_heatmap
```
__Figure 4. Heatmap of differentially expressed genes.__ The heatmap includes differentially expressed genes. The same clustering seen in the MDS plot (Figure 1) is seen in the heatmap.

---

# Thresholded over-representation analysis
For the thresholded over-representation analysis, we'll be using the R interface for g:Profiler using the gprofiler2 package [@gprofiler]. For the annotation source, we will use GO biological process. This annotation source was chosen as it follows the process from the original study [@paper]. g:Profiler was chosen as the method 

For the thresholds, for upregulated, the p- value is less than 0.05 and the log fold change is greater than 0. For downregulated, the p- value is less than 0.05 and the log fold change is less than 0. These were determined in the previous section.

We will first see how many upregulated genes, downregulated genes, and differentially expressed genes there are. 
```{r}
upregulated_MB468 <- row.names(qlf_output_hits_MB468)[(which(qlf_output_hits_MB468$expression ==
                                                           "upregulated"))]
upregulated_R8 <- row.names(qlf_output_hits_R8)[(which(qlf_output_hits_R8$expression ==
                                                           "upregulated"))]
upregulated <- union(upregulated_MB468, upregulated_R8)
downregulated_MB468 <- row.names(qlf_output_hits_MB468)[(which(qlf_output_hits_MB468$expression ==
                                                           "downregulated"))]
downregulated_R8 <- row.names(qlf_output_hits_R8)[(which(qlf_output_hits_R8$expression ==
                                                           "downregulated"))]
downregulated <- union(downregulated_MB468, downregulated_R8)
all_degs <- union(upregulated, downregulated)

# number of upregulated genes
length(upregulated)

# number of downregulated genes
length(downregulated)

# number of differentially expressed genes
length(all_degs)
```


## Upregulated genes
With the upregulated genes, no terms were found to be significantly enriched. 
```{r, message = FALSE, warning = FALSE}
upreg_res <- gost(upregulated, organism = "hsapiens", sources = "GO:BP")

if(is.null(upreg_res)) {
  print("No significantly enriched terms")
}
```



## Downregulated genes
With the downregulated genes, two terms were found to be significantly enriched. 
```{r}
downreg_res <- gost(downregulated, organism = "hsapiens", sources = "GO:BP")

downreg_df <- data.frame(term_id = downreg_res$result$term_id, 
                      term_name = downreg_res$result$term_name, 
                      pval = downreg_res$result$p_value)

kable(downreg_df, caption = "Table 5. Significantly Enriched Terms (Downregulated Genes)", col.names = c("Term ID", "Term Names", "P-value")) %>%
  kable_styling()
```



## All differentially expressed genes
With the all differentialy expressed genes, two terms were found to be significantly enriched. 
```{r}
all_degs_res <- gost(all_degs, organism = "hsapiens", sources = "GO:BP")

all_degs_df <- data.frame(term_id = all_degs_res$result$term_id, 
                      term_name = all_degs_res$result$term_name, 
                      pval = all_degs_res$result$p_value)

kable(all_degs_df, caption = "Table 6. Significantly Enriched Terms (All Differentially Expressed Genes)", col.names = c("Term ID", "Term Names", "P-value")) %>%
  kable_styling()
```


# Interpretation
Three terms that were significantly enriched terms were related to response to stimulus. Given that the samples' environment changed when methionine stress was added, these pathways were likely enriched because the cell lines were responding to the change in environment.

Four of the terms that were significantly enriched terms were related to cell death. Cancer cells have a metabolic requirement for methionine [@paper]. Therefore, when methionine was replaced with homocysteine, the TNBC cells were deprived of an essential metabolic requirement, possibly leading to cell death. Therefore, these pathways relating to cell death were significantly enriched.

Two significantly enriched terms were response to corticotropin-releasing hormone. There are a few studies that observe some link between corticotropin-releasing hormone and methionine. One study showed that peptide histidine methionine strongly potentiated corticotropin-releasing hormone induced adrenocorticotropic hormone (ACTH) secretion in humans [@mouse]. However, the exact link between these pathways and the experimental data is unclear.


# References
@paper
@edgeR
@limma
@ComplexHeatmap
@circlize
@knitr
@kableExtra
@ggplot
@dplyr
@gprofiler
@mouse